---
title: "FuDGE"
output:
  html_document: default
  pdf_document: default
date: "2023-08-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Setup

```{r}
require(phytools)
require(picante)
require(tidyr)
require(dplyr)
setwd("~/Dropbox/FuDGE/data_for_submission/")
```

## 2. Loading and Preparing the Trees

We begin by loading the function-derived phylogram from Skeels and Yaxley (2023), which represents functional relationships between bird species based on morphological traits.

```{r}
fun_tree <- read.tree('UPGMA_functional_tree_birds.tree')
plot(fun_tree, show.tip.label = F)
```

We also load a molecular time-calibrated consensus phylogeny from Jetz et al. (2012).

```{r}
phy <- read.tree('singe_bird_phylo.tre')
plot(phy, show.tip.label = F)
```

To make our measures of functional and evolutionary distinctiveness directly comparable, we rescale the functional tree so that its total depth matches that of the molecular phylogeny.

```{r}
require(geiger)
fun_tree <- rescale(fun_tree, model = 'depth', depth = max(nodeHeights(phy)))
c(max(nodeHeights(phy)), max(nodeHeights(fun_tree)))
```

## 3. Computing Functional and Evolutionary Distinctiveness

We calculate Functional Distinctiveness (FuD) scores using the `evol.distinct()` function from the `picante` package, applied to the functional tree.

```{r}
FuD <- evol.distinct(fun_tree)
head(FuD)
```

Let's examine the distribution of these scores.

```{r}
hist(FuD$w)
```

The distribution shows considerable right skew. A Shapiro-Wilk test confirms it is not normally distributed.

```{r}
shapiro.test(FuD$w[sample(1:nrow(FuD), 5000)])
```

We calculate Evolutionary Distinctiveness (ED) scores using the same procedure, this time applied to the molecular phylogeny.

```{r}
ED <- evol.distinct(phy)
hist(ED$w)
```

Again we have considerable right skew. This is typical for these indices; the standard solution is to take the natural log of the values before analysis.

## 4. Combining ED and FuD Scores

It is possible that some species appear on one tree but not the other. Let's check for mismatches.

```{r}
c(length(which(ED$Species %in% FuD$Species == F)), length(which(FuD$Species %in% ED$Species == F)))
```

We have 1,919 species on the molecular phylogeny for which we lack morphological data. We combine the two score sets into a single data frame, assigning `NA` to FuD scores for those missing species.

```{r}
df <- tibble(Species = ED$Species, ED = ED$w) %>% arrange(Species)
df$FuD <- NA
FuD <- FuD %>% arrange(Species)
df <- df %>% arrange(Species)
df$FuD[which(df$Species %in% FuD$Species)] <- FuD$w
head(df)
```

## 5. Exploring the ED–FuD Relationship

### 5.1 Spearman's rank correlation

Since neither index is normally distributed, we use Spearman's rank correlation — a non-parametric test.

```{r}
cor.test(df$ED, df$FuD, method = 'spearman')
```

A rho of 0.341 is moderate evidence of a positive correlation. Let's inspect this visually using log-transformed axes to spread out the heavily right-skewed distributions.

```{r}
require(ggplot2)

ggplot(df, aes(x = ED, y = FuD)) +
  geom_point() +
  geom_smooth(method = 'lm', se = T) +
  scale_y_continuous(trans = "log10") +
  scale_x_continuous(trans = "log10") +
  theme_classic()
```

### 5.2 Overlap between top 100 species

Let's compare the top 100 species ranked by FuD...

```{r}
df <- df[order(df$FuD, decreasing = T),]
FuD_Species <- df$Species[1:100]
df[1:100,]
```

...and by ED.

```{r}
df <- df[order(df$ED, decreasing = T),]
ED_Species <- df$Species[1:100]
df[1:100,]
```

```{r}
overlap <- FuD_Species[which(FuD_Species %in% ED_Species == T)]
overlap
```

Very little overlap — only ten species are common to both lists. This suggests that ED and FuD identify fundamentally different sets of priority species.

## 6. Phylogenetic Generalised Least Squares (PGLS) Analysis

### 6.1 Preparing data for PGLS

We expect both ED and FuD to be phylogenetically non-independent: closely related species will tend to have similar scores. To test whether ED predicts FuD above and beyond the correlation generated by shared evolutionary history, we need to account for phylogenetic structure using Phylogenetic Generalised Least Squares (PGLS).

```{r}
library(geiger)
library(nlme)

ED <- unlist(as.vector(df[,'ED']))
names(ED) <- as.vector(df$Species)

FuD <- unlist(as.vector(df[,'FuD']))
names(FuD) <- as.vector(df$Species)
```

We remove species missing FuD scores and trim the phylogeny accordingly.

```{r}
complete_tips <- names(which(is.na(FuD) == F))
FuD <- FuD[complete_tips]
ED <- ED[complete_tips]
```

```{r}
phy_trim <- keep.tip(phy, complete_tips)
```

### 6.2 Fitting the global models

```{r}
df$log_ED  <- log(df$ED)
df$log_FuD <- log(df$FuD)


# this is extremely slow so feel free just to load the pglsModel object using
# pglsModel <- gls(log_FuD ~ log_ED, correlation = corBrownian(1, phy = phy_trim, form = ~ Species),
#      data = na.omit(df), method = "ML")
# save(pglsModel, 'pglsModel.Rdata')

# Pre-fitted model loaded from disk to avoid long computation time
load('pglsModel.Rdata')
summary(pglsModel)
```

The PGLS detects no statistically significant association between the two variables. An increase of ln(ED) by one corresponds to an increase of only ~0.04 in ln(FuD). We can express this as a multiplicative percentage change.

```{r}
setNames((c(exp(1), exp(pglsModel$coefficients[[2]]))-1)*100,
nm = c('ED increase %', 'FuD increase %'))
```

A doubling of ED (~172% increase) equates to only a ~4.3% increase in FuD. For comparison, here is the equivalent GLS without a phylogenetic correction.

```{r}
glsModel <- gls(log_FuD ~ log_ED, data = na.omit(df), method = "ML")
summary(glsModel)
```

```{r}
setNames((c(exp(1), exp(glsModel$coefficients[[2]]))-1)*100,
nm = c('ED increase %', 'FuD increase %'))
```

Without the phylogenetic correction, a doubling of ED predicts a ~19% increase in FuD — a stronger but still modest relationship.

```{r}
plot(
  df$log_ED, df$log_FuD,
  xlab = "ln(ED)",
  ylab = "ln(FuD)",
  pch  = 16
)
abline(a = coef(pglsModel)[1], b = coef(pglsModel)[2], col = "red", lwd = 2)
abline(a = coef(glsModel)[1],  b = coef(glsModel)[2],  col = "blue", lwd = 2)
legend("topleft", legend = c("PGLS", "GLS"), col = c("red", "blue"), lwd = 2, bty = "n")
```

The blue line (GLS) is steeper than the red line (PGLS), confirming that much of the apparent relationship between ED and FuD is explained by the phylogenetic covariance of both indices.

## 7. Clade-Level Analysis

### 7.1 Adding taxonomic information

We now ask how the ED–FuD relationship varies across bird orders. We load the AVONET database (Tobias et al. 2022) to obtain order-level taxonomy.

```{r}
AVONET <- read.csv('AVONET1_BirdLife.csv')
AVONET <- AVONET[,2:4]
head(AVONET)
```

```{r}
AVONET$Species1 <- gsub(" ", "_", AVONET$Species1)
AVONET <- AVONET[which(AVONET$Species1 %in% df$Species == T),]
df <- na.omit(df)
df <- df[order(df$Species),]
AVONET <- AVONET[order(AVONET$Species1),]
AVONET <- AVONET[-which(AVONET$Species1 %in% df$Species == F),]

any(df$Species != AVONET$Species1)

df$Family <- AVONET$Family1
df$Order  <- AVONET$Order1
```

```{r}
unique(df$Order)
```

```{r}
df$Order <- as.factor(df$Order)
table(df$Order)


```

### 7.2 Creating order dummy variables

We create a binary dummy variable for each order to allow interaction terms in our model.

```{r}
require(caret)
dummy_model <- dummyVars(~ Order, data = df)
dummy_data  <- predict(dummy_model, newdata = df)
df <- bind_cols(df, dummy_data)
```

To avoid the dummy variable trap (perfect multicollinearity), we drop the Passeriformes dummy, making passerines the reference level. All other order coefficients will therefore represent departures from the passerine baseline.

```{r}
df$Order.Passeriformes <- NULL
```

### 7.3 Building the interaction formula

We construct a model formula with an interaction term between log(ED) and each order dummy, allowing us to estimate how the slope of the FuD–ED relationship varies between orders.

```{r}
df$sc.log_ED  <- scale(df$log_ED)
df$sc.log_FuD <- scale(df$log_FuD)

orders <- unique(df$Order)
orders <- orders[which(orders != 'Passeriformes')]
interactions <- paste0('log_ED:', paste0('Order.', orders))
interactions <- paste(interactions, collapse = " + ")

orders <- paste(paste0('Order.', orders), collapse = " + ")

fml <- c('log_FuD ~ log_ED + ', orders, '+', interactions)
fml <- paste(fml, collapse = '')
fml <- as.formula(fml)
fml
```

### 7.4 GLS clade model (without phylogenetic correction)

```{r}
glsModel_clade <- gls(model = fml,
    data = na.omit(df), method = "ML",
    control = list(singular.ok = TRUE))
```

The per-order slope is the sum of the global log_ED coefficient (the Passerine baseline) and the relevant interaction term. We verify this against a reduced model fit to Podicipediformes only.

```{r}
coefficents <- coef(glsModel_clade)
coefficents['log_ED'] + coefficents['log_ED:Order.Podicipediformes']
```

```{r}
glsModel_pass <- gls(model = log_FuD ~ log_ED,
    data = na.omit(df[which(df$Order.Podicipediformes == 1),]), method = "ML",
    control = list(singular.ok = TRUE))
coef(glsModel_pass)[2]
```

The two estimates match, confirming the interaction approach is working correctly.

```{r}
coefficients <- coef(glsModel_clade)
conf95  <- confint(glsModel_clade)
summary <- summary(glsModel_clade)
coefficents <- tibble(Pred = names(coefficients), Coef = coefficients, low = conf95[,1],
                      high = conf95[,2], p_value = summary$tTable[,4])

order_coefs <- coefficents[grep(':Order.', coefficents$Pred),]

order_coefs$Coef <- order_coefs$Coef + coefficents[[2,2]]
order_coefs$low  <- order_coefs$low  + coefficents[[2,3]]
order_coefs$high <- order_coefs$high + coefficents[[2,4]]

new_row <- tibble(Pred = 'Passeriformes',
                  Coef = coefficents[[2,2]],
                  low  = conf95[[2,1]],
                  high = conf95[[2,2]],
                  p_value = summary$tTable[2,4])

order_coefs <- bind_rows(order_coefs, new_row)
order_coefs$Pred <- gsub('log_ED:Order.', '', order_coefs$Pred)
order_coefs$sig  <- ifelse(order_coefs$p_value < 0.05, T, F)
write.csv(order_coefs, 'order_coefsGLS.csv')

```

```{r}
g3 <- ggplot(order_coefs, aes(x = Pred, y = Coef, col = sig)) +
  geom_point() +
  geom_linerange(aes(xmin = Pred, xmax = Pred, ymin = low, ymax = high)) +
  geom_hline(yintercept = 0, lty = 2) +
  theme_classic() +
  ylab('Slope of FuD ~ ED') +
  xlab("") +
  scale_color_manual(values = c('darkgrey', 'black')) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = 'none') +
  geom_vline(xintercept = unique(order_coefs$Pred), color = 'gray', size = .1)

g3
```

Most orders show a slight positive relationship between FuD and ED; those with negative slopes are insignificant with wide confidence intervals. Let's add species counts to understand whether these orders are simply data-poor.

```{r}
order_counts <- table(na.omit(df)$Order)
order_coefs$n <- c(order_counts[match(order_coefs$Pred, names(order_counts))])

g4 <- ggplot(order_coefs, aes(x = Pred, y = Coef, col = sig, label = n)) +
  geom_point() +
  geom_linerange(aes(xmin = Pred, xmax = Pred, ymin = low, ymax = high)) +
  geom_hline(yintercept = 0, lty = 2) +
  geom_text(aes(y = -Inf), position = position_nudge(y = 0.1), vjust = -0.5, size = 2.2, col = 'blue') +
  theme_classic() +
  ylab('Slope of FuD ~ ED') +
  xlab("") +
  scale_color_manual(values = c('darkgrey', 'black')) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = 'none') +
  geom_vline(xintercept = unique(order_coefs$Pred), color = 'gray', size = .1)

ggsave('~/Dropbox/FuDGE/Figures/order_gls.tiff', g4, width = 7, height = 4)

g4
```

As expected, the orders with negative coefficients tend to have very few species. The only order with a reasonably large sample size and a negative coefficient is the Podicipediformes (grebes), though even here the relationship is not significant.

The only problem is that we are treating our FuD scores as if they are independent of one another, but they are not — species are evolved units with shared history. Closely related species will tend to have similar ED scores, and possibly similar FuD scores too, as being functionally unremarkable may be an inherent feature of some clades. If this is true, any correlation between ED and FuD may partly reflect shared history rather than a direct relationship between the two. We therefore fit the equivalent model with a phylogenetic correction.

### 7.5 PGLS clade model (with phylogenetic correction)

```{r}
# again this is slow so feel free to just load the final model 
#pglsModel_clade <- gls(model = fml, data = na.omit(df), method = "ML", correlation = corBrownian(1, phy = phy_trim, form = ~ Species), control = list(singular.ok = TRUE))


# Pre-fitted model loaded from disk to avoid long computation time
pglsModel_clade <- readRDS('pglsModel_clade.rds')
summary <- summary(pglsModel_clade)
```

```{r}
order_coefs_nocorrection <- order_coefs
coefficients <- coef(pglsModel_clade)
conf95  <- confint(pglsModel_clade)
coefficents <- tibble(Pred = names(coefficients), Coef = coefficients, low = conf95[,1],
                      high = conf95[,2], p_value = summary$tTable[,4])

order_coefs <- coefficents[grep(':Order.', coefficents$Pred),]
order_coefs$Coef <- order_coefs$Coef + coefficents[[2,2]]
order_coefs$low  <- order_coefs$low  + coefficents[[2,3]]
order_coefs$high <- order_coefs$high + coefficents[[2,4]]


new_row <- tibble(Pred = 'Passeriformes',
                  Coef = coefficents[[2,2]],
                  low  = conf95[[2,1]],
                  high = conf95[[2,2]],
                  p_value = summary$tTable[2,4])

order_coefs <- bind_rows(order_coefs, new_row)
order_coefs$Pred <- gsub('log_ED:Order.', '', order_coefs$Pred)
order_coefs$sig  <- ifelse(order_coefs$p_value < 0.05, T, F)
order_counts <- table(na.omit(df)$Order)
order_coefs$Pred <- gsub('sc.', '', order_coefs$Pred)
order_coefs$n <- c(order_counts[match(order_coefs$Pred, names(order_counts))])

write.csv(order_coefs, 'order_coefsPGLS.csv')

g5 <- ggplot(order_coefs, aes(x = Pred, y = Coef, col = sig, label = n)) +
  geom_point() +
  geom_linerange(aes(xmin = Pred, xmax = Pred, ymin = low, ymax = high)) +
  geom_hline(yintercept = 0, lty = 2) +
  geom_text(aes(y = -Inf), position = position_nudge(y = 0.1), vjust = -0.5, size = 2.2, col = 'blue') +
  theme_classic() +
  ylab('Slope of FuD ~ ED') +
  xlab("") +
  scale_color_manual(values = c('darkgrey', 'black')) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = 'none') +
  geom_vline(xintercept = unique(order_coefs$Pred), color = 'gray', size = .1)

ggsave('~/Dropbox/FuDGE/Figures/order_pgls.tiff', g5, width = 7, height = 4)

g5
```

Including the phylogenetic correction pulls all coefficients towards zero and renders them non-significant, suggesting that much of the variation in FuD can be explained by shared evolutionary history without invoking ED.

### 7.6 Figure 1: Combined coefficient plot and scatter

```{r}
order_coefs$model <- 'PGLS'
order_coefs_nocorrection$model <- 'GLS'
all_coefs <- rbind(order_coefs, order_coefs_nocorrection)

library(ggh4x)
all_coefs$model <- factor(all_coefs$model, levels = c("GLS", "PGLS"))

all_coefs$Pred <- gsub("sc.", "",all_coefs$Pred)

F1.2 <- ggplot(all_coefs, aes(x = Pred, y = Coef, col = sig, label = n)) +
  geom_point() +
  geom_errorbar(aes(ymin = low, ymax = high), width = 0) +
  geom_hline(yintercept = 0, lty = 2) +
  geom_text(
    data = subset(all_coefs, model == "PGLS"),
    aes(label = n, y = -Inf),
    position = position_nudge(y = 0.1),
    vjust = -0.5,
    size = 2.2,
    colour = "black"
  ) +
  theme_classic() +
  ylab("Slope of ln(FuD) ~ ln(ED)") +
  xlab("") +
  scale_color_manual(values = c("darkgrey", "black")) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    legend.position = "none",
    strip.text = element_text(color = "white", face = "bold")
  ) +
  geom_vline(xintercept = unique(all_coefs$Pred), colour = "gray", linewidth = .1, alpha = 0.5) +
  facet_wrap2(
    ~model, ncol = 1, scales = "free_y",
    strip = strip_themed(
      background_x = elem_list_rect(fill = c("#3f6faf", "#f49843"))
    )
  )
F1.2
```

```{r}
line_df <- data.frame(
  model     = c("PGLS", "GLS"),
  intercept = c(coef(pglsModel)[1], coef(glsModel)[1]),
  slope     = c(coef(pglsModel)[2], coef(glsModel)[2])
)

F1.1 <- ggplot(df, aes(x = log_ED, y = log_FuD)) +
  geom_point(shape = 16, alpha = 0.5) +
  geom_abline(
    data = line_df,
    aes(intercept = intercept, slope = slope, colour = model),
    linewidth = 1.2
  ) +
  labs(x = "ln(ED)", y = "ln(FuD)", colour = "Model") +
  scale_colour_manual(values = c("PGLS" = "#f49843", "GLS" = "#3f6faf")) +
  theme_bw() +
  coord_equal(xlim = c(-2, 4), ylim = c(-2, 4)) +
  theme(
    legend.position  = c(0.05, 0.95),
    legend.justification = c(0, 1)
  )
F1.1
```

```{r}
require(ggpubr)
F1 <- ggarrange(F1.1, F1.2, ncol = 2, nrow = 1, widths = c(2,3), labels = c('A', 'B'))
F1
ggsave('~/Dropbox/FuDGE/Figures/Figure1.pdf', plot = F1, width = 180, height = 70, units = 'mm', scale = 1.7)
```

### 7.7 Pseudo-R² and supplementary figure

We calculate pseudo-R² for each model as the squared correlation between fitted and observed values.

```{r}
R2GLS      <- cor(df$log_FuD, fitted(glsModel))^2
R2PGLS     <- cor(df$log_FuD, fitted(pglsModel))^2
R2GLSclade <- cor(df$log_FuD, fitted(glsModel_clade))^2
R2PGLSclade <- cor(df$log_FuD, fitted(pglsModel_clade))^2
```

```{r}
order_df <- df %>%
  group_by(Order) %>%
  summarise(ED_m = mean(ED), ED_sd = sd(ED), FuD_m = mean(FuD), FuD_sd = sd(FuD)) %>%
  pivot_longer(
    cols = c(ED_m, FuD_m, ED_sd, FuD_sd),
    names_to = c("metric", ".value"),
    names_pattern = "(ED|FuD)_(m|sd)"
  )

FigS1 <- ggplot(order_df, aes(x = m, y = Order, colour = metric)) +
  geom_point(position = position_dodge(width = 0.6), size = 3) +
  geom_errorbarh(
    aes(xmin = m - sd, xmax = m + sd),
    height = 0.2,
    position = position_dodge(width = 0.6)
  ) +
  scale_colour_manual(
    values = c("ED" = "#f49843", "FuD" = "#3f6faf"),
    labels = c("ED", "FuD")
  ) +
  labs(x = "Mean value", y = NULL, colour = NULL) +
  theme_bw() +
  theme(legend.position = "top") +
  scale_y_discrete(limits = rev)

ggsave('~/Dropbox/FuDGE/Figures/FigureS1.pdf', plot = FigS1, width = 150, height = 180, units = 'mm', scale = 1.3)

FigS1
```

## 8. FuDGE and EDGE as Conservation Prioritisation Indices

### 8.1 Loading IUCN threat data

We load the AVONET taxonomy crosswalk and IUCN Red List assessments.

```{r}
crossWalk <- read.csv('BirdLife-BirdTree crosswalk.csv')
head(crossWalk)
```

```{r}
crossWalk$Species3 <- gsub(" ", "_", crossWalk$Species3)
crossWalk$Species1 <- gsub(" ", "_", crossWalk$Species1)
```

```{r}
iucn <- read.csv('assessments.csv')
head(iucn)[,1:5]
```

```{r}
iucn$scientificName <- gsub(" ", "_", iucn$scientificName)
c(length(which(df$Species %in% iucn$scientificName == F)),
  length(which(crossWalk$Species1 %in% iucn$scientificName == F)),
  length(which(crossWalk$Species3 %in% iucn$scientificName == F)))
```

We have good name coverage using the species names in `df`. We add the IUCN categories and remove species with missing or uninformative assessments.

```{r}
df$assessment <- NA
df$assessment <- iucn$redlistCategory[match(df$Species, iucn$scientificName)]
df$assessment <- as.factor(df$assessment)
head(df)
```

```{r}
df <- na.omit(df)
df <- df[-which(df$assessment == 'Data Deficient'),]
df <- df[-which(df$assessment == 'Extinct'),]
df <- df[-which(df$assessment == 'Extinct in the Wild'),]

unique(df$assessment)
```

### 8.2 Creating the Global Endangerment (GE) score

We encode IUCN categories as an ordered numeric variable (0–4) for use in the DGE formula.

```{r}
df$GE <- NA
df$GE[which(df$assessment == 'Least Concern')]        <- 0
df$GE[which(df$assessment == 'Near Threatened')]      <- 1
df$GE[which(df$assessment == 'Vulnerable')]           <- 2
df$GE[which(df$assessment == 'Endangered')]           <- 3
df$GE[which(df$assessment == 'Critically Endangered')] <- 4
hist(df$GE)
```

### 8.3 Calculating EDGE and FuDGE scores

We define the DGE function, where D is a distinctiveness measure (ED or FuD) and GE is the global endangerment score.

```{r}
DGE <- function(D, GE) log(1 + D) + GE * log(2)
```

```{r}
df$FuDGE <- DGE(D = df$FuD, GE = df$GE)
df$EDGE  <- DGE(D = df$ED,  GE = df$GE)
```


```{r}
write.csv(df,'complete_data.csv')
```



### 8.5 Overlap among priority species lists

We extract the top 100 species under each of the four indices and compare them pairwise.

```{r}
ED_species     <- df$Species[order(df$ED,    decreasing = T)][1:100]
FuD_species    <- df$Species[order(df$FuD,   decreasing = T)][1:100]
EDGE_species   <- df$Species[order(df$EDGE,  decreasing = T)][1:100]
ecoDGE_species <- df$Species[order(df$FuDGE, decreasing = T)][1:100]

priority_lists <- setNames(list(ED_species, FuD_species, EDGE_species, ecoDGE_species),
                           c('ED Priorities', 'FuD Priorities', 'EDGE Priorities', 'ecoDGE Priorities'))
```

```{r}
num_lists <- length(priority_lists)

pairwise_matrix <- matrix(0, ncol = num_lists, nrow = num_lists)
colnames(pairwise_matrix) <- rownames(pairwise_matrix) <- paste("List", 1:num_lists)

for (i in 1:num_lists) {
  for (j in 1:num_lists) {
    shared_species <- length(intersect(priority_lists[[i]], priority_lists[[j]]))
    pairwise_matrix[i, j] <- shared_species
  }
}

rownames(pairwise_matrix) <- colnames(pairwise_matrix) <- names(priority_lists)
print(pairwise_matrix)
```

```{r}
comparison_data <- as.data.frame(as.table(pairwise_matrix))
colnames(comparison_data) <- c("List1", "List2", "SharedSpecies")

comparison_data_lower <- comparison_data %>%
  filter(as.numeric(List1) >= as.numeric(List2))

heatmap_plot <- ggplot(comparison_data_lower, aes(x = List1, y = List2, fill = SharedSpecies)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "lightblue") +
  geom_text(aes(label = SharedSpecies), color = "black", size = 3) +
  labs(title = "", x = "", y = "", fill = "N shared species") +
  theme_minimal() +
  coord_equal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = 'bottom')

print(heatmap_plot)
```

```{r}
DGE_species <- EDGE_species[EDGE_species %in% ecoDGE_species]

DGE_species <- df %>%
  filter(Species %in% DGE_species)

table(DGE_species$assessment)

DGE_species %>%
  select(Species, assessment)
```

All four indices identify quite different sets of priority species. EDGE and ecoDGE are the most aligned but still share fewer than half their top 100 species. This surprisingly low agreement suggests the two indices are measuring different aspects of biodiversity. While we would not expect raw ED or FuD alone to guide conservation priorities, the low overlap between EDGE and ecoDGE calls for a direct test of how effectively each index protects the diversity it is designed to represent.

## 9. Extinction Simulations

### 9.1 Assigning extinction probabilities

We translate IUCN categories into extinction probabilities following Davis et al. (2018).

```{r}
df$extinction_prob <- NA
df$extinction_prob[which(df$assessment == 'Least Concern')]         <- 0.0017
df$extinction_prob[which(df$assessment == 'Near Threatened')]       <- 0.0141
df$extinction_prob[which(df$assessment == 'Vulnerable')]            <- 0.1
df$extinction_prob[which(df$assessment == 'Endangered')]            <- 0.6723
df$extinction_prob[which(df$assessment == 'Critically Endangered')] <- 0.999
```

### 9.2 Stochastic extinction function

```{r}
extinction_roll <- function(p) {
  e = runif(1, 0, 1)
  if (e < p) {
    return('extinct')
  } else {
    return('survived')
  }
}
set.seed(888)
c(extinction_roll(.5), extinction_roll(.5), extinction_roll(.5), extinction_roll(.5))
```

### 9.3 Single-run scenario (optimistic)

```{r}
set.seed(888)
keepers <- df$Species[which(unlist((lapply(df$extinction_prob, extinction_roll))) == 'survived')]
phy <- keep.tip(phy, df$Species)
phy_extinct <- keep.tip(phy, keepers)

1 - (sum(phy_extinct$edge.length) / sum(phy$edge.length))
```

Approximately 3% of phylogenetic diversity is lost under the optimistic scenario. Note that this figure is based on a reduced phylogeny because our morphological dataset does not cover all species in the Jetz tree.

```{r}
fun_tree <- keep.tip(fun_tree, df$Species)
fun_extinct <- keep.tip(fun_tree, keepers)

1 - (sum(fun_extinct$edge.length) / sum(fun_tree$edge.length))
```

About 5% of functional diversity is lost. For comparison, we shuffle extinction probabilities randomly across species.

```{r}
random_endangerment <- sample(df$extinction_prob, nrow(df), replace = F)
keepers <- df$Species[which(unlist((lapply(random_endangerment, extinction_roll))) == 'survived')]
phy_extinct <- keep.tip(phy, keepers)

1 - (sum(phy_extinct$edge.length) / sum(phy$edge.length))
```

```{r}
fun_extinct <- keep.tip(fun_tree, keepers)
1 - (sum(fun_extinct$edge.length) / sum(fun_tree$edge.length))
```

### 9.4 Single-run scenario (pessimistic)

```{r}
df$extinction_prob[which(df$assessment == 'Least Concern')]   <- 0.01269
df$extinction_prob[which(df$assessment == 'Near Threatened')] <- 0.09
df$extinction_prob[which(df$assessment == 'Vulnerable')]      <- 0.605
df$extinction_prob[which(df$assessment == 'Endangered')]      <- 0.8991
```

```{r}
keepers <- df$Species[which(unlist((lapply(df$extinction_prob, extinction_roll))) == 'survived')]
phy_extinct <- keep.tip(phy, keepers)
1 - (sum(phy_extinct$edge.length) / sum(phy$edge.length))
```

Under the pessimistic scenario we lose about 7% of PD...

```{r}
fun_extinct <- keep.tip(fun_tree, keepers)
1 - (sum(fun_extinct$edge.length) / sum(fun_tree$edge.length))
```

...and around 10% of FD.

```{r}
random_endangerment <- sample(df$extinction_prob, nrow(df), replace = F)
keepers <- df$Species[which(unlist((lapply(random_endangerment, extinction_roll))) == 'survived')]
phy_extinct <- keep.tip(phy, keepers)
1 - (sum(phy_extinct$edge.length) / sum(phy$edge.length))
```

```{r}
fun_extinct <- keep.tip(fun_tree, keepers)
1 - (sum(fun_extinct$edge.length) / sum(fun_tree$edge.length))
```

### 9.5 Stochastic simulation (1,000 iterations)

Because extinction is a stochastic process, we repeat each scenario 1,000 times to characterise average expected loss. Note: this loop may take some time to run.

```{r}
PDlost     <- vector()
FDlost     <- vector()
PDlost_rnd <- vector()
FDlost_rnd <- vector()

for(i in 1:1000) {
  keepers <- df$Species[which(unlist((lapply(df$extinction_prob, extinction_roll))) == 'survived')]
  phy_extinct <- keep.tip(phy, keepers)
  PDlost <- c(PDlost, 1 - (sum(phy_extinct$edge.length) / sum(phy$edge.length)))
  fun_extinct <- keep.tip(fun_tree, keepers)
  FDlost <- c(FDlost, 1 - (sum(fun_extinct$edge.length) / sum(fun_tree$edge.length)))
  random_endangerment <- sample(df$extinction_prob, nrow(df), replace = F)
  keepers <- df$Species[which(unlist((lapply(random_endangerment, extinction_roll))) == 'survived')]
  phy_extinct <- keep.tip(phy, keepers)
  PDlost_rnd <- c(PDlost_rnd, 1 - (sum(phy_extinct$edge.length) / sum(phy$edge.length)))
  fun_extinct <- keep.tip(fun_tree, keepers)
  FDlost_rnd <- c(FDlost_rnd, 1 - (sum(fun_extinct$edge.length) / sum(fun_tree$edge.length)))
}

output <- tibble(PDlost, FDlost, PDlost_rnd, FDlost_rnd, run = 1:length(PDlost), treatment = 'pessimistic')
output
```

We repeat with the optimistic extinction probabilities.

```{r}
df$extinction_prob[which(df$assessment == 'Least Concern')]         <- 0.0017
df$extinction_prob[which(df$assessment == 'Near Threatened')]       <- 0.0141
df$extinction_prob[which(df$assessment == 'Vulnerable')]            <- 0.1
df$extinction_prob[which(df$assessment == 'Endangered')]            <- 0.6723
df$extinction_prob[which(df$assessment == 'Critically Endangered')] <- 0.999
df$extinction_prob[which(df$assessment == 'Extinct in the Wild')]   <- 1
df$extinction_prob[which(df$assessment == 'Extinct')]               <- 1

PDlost     <- vector()
FDlost     <- vector()
PDlost_rnd <- vector()
FDlost_rnd <- vector()

for(i in 1:1000) {
  keepers <- df$Species[which(unlist((lapply(df$extinction_prob, extinction_roll))) == 'survived')]
  phy_extinct <- keep.tip(phy, keepers)
  PDlost <- c(PDlost, 1 - (sum(phy_extinct$edge.length) / sum(phy$edge.length)))
  fun_extinct <- keep.tip(fun_tree, keepers)
  FDlost <- c(FDlost, 1 - (sum(fun_extinct$edge.length) / sum(fun_tree$edge.length)))
  random_endangerment <- sample(df$extinction_prob, nrow(df), replace = F)
  keepers <- df$Species[which(unlist((lapply(random_endangerment, extinction_roll))) == 'survived')]
  phy_extinct <- keep.tip(phy, keepers)
  PDlost_rnd <- c(PDlost_rnd, 1 - (sum(phy_extinct$edge.length) / sum(phy$edge.length)))
  fun_extinct <- keep.tip(fun_tree, keepers)
  FDlost_rnd <- c(FDlost_rnd, 1 - (sum(fun_extinct$edge.length) / sum(fun_tree$edge.length)))
}

output2 <- tibble(PDlost, FDlost, PDlost_rnd, FDlost_rnd, run = 1001:2000, treatment = 'optimistic')
output <- bind_rows(output, output2)
output
```

We reshape to long format and add a variable distinguishing empirical from shuffled extinction rates.

```{r}
output <- gather(output, index, loss, PDlost:FDlost_rnd)
head(output)
```

```{r}
output$ER_treatment <- 'empirical'
output$ER_treatment[grep('_rnd', output$index)] <- 'random'
output$index <- gsub('_rnd', '', output$index)
```

```{r}
library(gghalves)

dodge <- position_dodge(width = 0.9)

g6 <- ggplot(output, aes(x = index, y = loss * 100, fill = ER_treatment)) +
  geom_violin(trim = FALSE, alpha = 0.8, position = dodge) +
 # geom_boxplot(width = 0.12, outlier.shape = NA, alpha = 0.6, position = dodge) +
  stat_summary(fun = median, geom = "point", position = dodge, size = 2, colour = "black") +
  facet_wrap(~treatment, ncol = 1, nrow = 2,
             labeller = labeller(
               treatment = c(optimistic = "Optimistic", pessimistic = "Pessimistic")
             )) +
  scale_fill_manual(values = c('#3f6faf', '#f49843'),
                    labels = c('empirical' = 'Empirical extinction',
                               'random' = 'Shuffled extinction')) +
  scale_x_discrete(labels = c("FDlost" = "FD", "PDlost" = "PD")) +
  labs(y = "Percentage diversity lost", x = NULL) +
  theme_bw() +
  theme(strip.text = element_text(face = "bold"),
        legend.title = element_blank(),
        legend.position = 'bottom')

g6
```

```{r}
output %>%
  group_by(index, treatment, ER_treatment) %>%
  summarise(average_loss = mean(loss)*100, sd_loss = sd(loss)*100)
```

FD appears to be more imperiled than PD under empirical extinction rates. We test this formally.

```{r}
require(broom)
output %>%
  group_by(treatment, index) %>%
  summarise(
    test = list(t.test(loss ~ ER_treatment)),
    .groups = "drop"
  ) %>%
  mutate(tidy = lapply(test, tidy)) %>%
  unnest(tidy)
```


## 10. Conservation Prioritisation Efficiency

We test how effectively each index preserves PD and FD when used to prioritise species for protection. Species are protected in rank order (by EDGE, FuDGE, or randomly), and their extinction probability is set to zero. We reload the phylogenies trimmed to only the species in our final dataset.

```{r}
EDGE_species  <- df$Species[order(df$EDGE,  decreasing = T)]
FuDGE_species <- df$Species[order(df$FuDGE, decreasing = T)]

df$extinction_prob[which(df$assessment == 'Least Concern')]         <- 0.0017
df$extinction_prob[which(df$assessment == 'Near Threatened')]       <- 0.0141
df$extinction_prob[which(df$assessment == 'Vulnerable')]            <- 0.1
df$extinction_prob[which(df$assessment == 'Endangered')]            <- 0.6723
df$extinction_prob[which(df$assessment == 'Critically Endangered')] <- 0.999

phy <- read.tree('singe_bird_phylo.tre')
phy <- keep.tip(phy = phy, tip = df$Species)
fun_tree <- read.tree('UPGMA_functional_tree_birds.tree')
fun_tree <- keep.tip(fun_tree, df$Species)
fun_tree <- rescale(fun_tree, model = 'depth', depth = max(nodeHeights(phy)))
```

### 10.1 Deterministic prioritisation curves (optimistic scenario)

We calculate PD and FD loss at intervals of 100 protected species.

```{r}
PDlost <- vector()
FDlost <- vector()
df_clone <- df

for(i in c(seq(0, nrow(df), 100), 8067)) {
  df_clone$extinction_prob[which(df_clone$Species %in% EDGE_species[1:i])] <- 0
  keepers <- df_clone$Species[which(unlist((lapply(df_clone$extinction_prob, extinction_roll))) == 'survived')]
  phy_extinct <- keep.tip(phy, keepers)
  PDlost <- c(PDlost, 1 - (sum(phy_extinct$edge.length) / sum(phy$edge.length)))
  fun_extinct <- keep.tip(fun_tree, keepers)
  FDlost <- c(FDlost, 1 - (sum(fun_extinct$edge.length) / sum(fun_tree$edge.length)))
}

output <- tibble(Runs = c(seq(0, nrow(df), 100), 8067), PD_EDGE = PDlost, FD_EDGE = FDlost)
tail(output)
```

```{r}
df_clone <- df

PDlost <- vector()
FDlost <- vector()

for(i in c(seq(0, nrow(df), 100), 8068)) {
  df_clone$extinction_prob[which(df_clone$Species %in% FuDGE_species[1:i])] <- 0
  keepers <- df_clone$Species[which(unlist((lapply(df_clone$extinction_prob, extinction_roll))) == 'survived')]
  phy_extinct <- keep.tip(phy, keepers)
  PDlost <- c(PDlost, 1 - (sum(phy_extinct$edge.length) / sum(phy$edge.length)))
  fun_extinct <- keep.tip(fun_tree, keepers)
  FDlost <- c(FDlost, 1 - (sum(fun_extinct$edge.length) / sum(fun_tree$edge.length)))
}

output$PD_FuDGE <- PDlost
output$FD_FuDGE <- FDlost
```

```{r}
random_species <- sample(df$Species, replace = F)
PDlost <- vector()
FDlost <- vector()
df_clone <- df

for(i in c(seq(0, nrow(df), 100), 8068)) {
  df_clone$extinction_prob[which(df_clone$Species %in% random_species[1:i])] <- 0
  keepers <- df_clone$Species[which(unlist((lapply(df_clone$extinction_prob, extinction_roll))) == 'survived')]
  phy_extinct <- keep.tip(phy, keepers)
  PDlost <- c(PDlost, 1 - (sum(phy_extinct$edge.length) / sum(phy$edge.length)))
  fun_extinct <- keep.tip(fun_tree, keepers)
  FDlost <- c(FDlost, 1 - (sum(fun_extinct$edge.length) / sum(fun_tree$edge.length)))
}

output$PD_Rnd <- PDlost
output$FD_Rnd <- FDlost
```

```{r}
output <- gather(output, index, loss, PD_EDGE:FD_Rnd)
head(output)
```

```{r}
output$treatment <- 'EDGE'
output$treatment[grep('_FuDGE', output$index)] <- 'ecoDGE'
output$index <- gsub('_EDGE',  '', output$index)
output$index <- gsub('_FuDGE', '', output$index)
output$treatment[grep('_Rnd', output$index)] <- 'Rnd'
output$index <- gsub('_Rnd', '', output$index)
head(output)
```

```{r}
g7 <- ggplot(output, aes(x = Runs, y = loss*100, col = treatment, lty = index)) +
  geom_line() +
  theme_classic() +
  ylab('Percentage lost') +
  xlab('Species protected') +
  scale_color_manual(values = c('blue', 'darkorange', 'darkgrey')) +
  scale_x_continuous(trans = 'log10') +
  theme(legend.title = element_blank())
g7
```

ecoDGE is better at protecting FD, as expected. Surprisingly, ecoDGE also does a good job at protecting PD, and by around 700–800 species protected it approaches EDGE's performance on PD while substantially outperforming it on FD.

### 10.2 Deterministic prioritisation curves (pessimistic scenario)

```{r}
df$extinction_prob[which(df$assessment == 'Least Concern')]   <- 0.01269
df$extinction_prob[which(df$assessment == 'Near Threatened')] <- 0.09
df$extinction_prob[which(df$assessment == 'Vulnerable')]      <- 0.605
df$extinction_prob[which(df$assessment == 'Endangered')]      <- 0.8991

PDlost <- vector()
FDlost <- vector()
df_clone <- df

for(i in c(seq(0, nrow(df), 100), 8067)) {
  df_clone$extinction_prob[which(df_clone$Species %in% EDGE_species[1:i])] <- 0
  keepers <- df_clone$Species[which(unlist((lapply(df_clone$extinction_prob, extinction_roll))) == 'survived')]
  phy_extinct <- keep.tip(phy, keepers)
  PDlost <- c(PDlost, 1 - (sum(phy_extinct$edge.length) / sum(phy$edge.length)))
  fun_extinct <- keep.tip(fun_tree, keepers)
  FDlost <- c(FDlost, 1 - (sum(fun_extinct$edge.length) / sum(fun_tree$edge.length)))
}

output <- tibble(Runs = c(seq(0, nrow(df), 100), 8067), PD_EDGE = PDlost, FD_EDGE = FDlost)
tail(output)

df_clone <- df

PDlost <- vector()
FDlost <- vector()

for(i in c(seq(0, nrow(df), 100), 8068)) {
  df_clone$extinction_prob[which(df_clone$Species %in% FuDGE_species[1:i])] <- 0
  keepers <- df_clone$Species[which(unlist((lapply(df_clone$extinction_prob, extinction_roll))) == 'survived')]
  phy_extinct <- keep.tip(phy, keepers)
  PDlost <- c(PDlost, 1 - (sum(phy_extinct$edge.length) / sum(phy$edge.length)))
  fun_extinct <- keep.tip(fun_tree, keepers)
  FDlost <- c(FDlost, 1 - (sum(fun_extinct$edge.length) / sum(fun_tree$edge.length)))
}

output$PD_FuDGE <- PDlost
output$FD_FuDGE <- FDlost

random_species <- sample(df$Species, replace = F)
PDlost <- vector()
FDlost <- vector()
df_clone <- df

for(i in c(seq(0, nrow(df), 100), 8068)) {
  df_clone$extinction_prob[which(df_clone$Species %in% random_species[1:i])] <- 0
  keepers <- df_clone$Species[which(unlist((lapply(df_clone$extinction_prob, extinction_roll))) == 'survived')]
  phy_extinct <- keep.tip(phy, keepers)
  PDlost <- c(PDlost, 1 - (sum(phy_extinct$edge.length) / sum(phy$edge.length)))
  fun_extinct <- keep.tip(fun_tree, keepers)
  FDlost <- c(FDlost, 1 - (sum(fun_extinct$edge.length) / sum(fun_tree$edge.length)))
}

output$PD_Rnd <- PDlost
output$FD_Rnd <- FDlost

output <- gather(output, index, loss, PD_EDGE:FD_Rnd)

output$treatment <- 'EDGE'
output$treatment[grep('_FuDGE', output$index)] <- 'ecoDGE'
output$index <- gsub('_EDGE',  '', output$index)
output$index <- gsub('_FuDGE', '', output$index)
output$treatment[grep('_Rnd', output$index)] <- 'Rnd'
output$index <- gsub('_Rnd', '', output$index)

g8 <- ggplot(output, aes(x = Runs, y = loss*100, col = treatment, lty = index)) +
  geom_line() +
  theme_classic() +
  ylab('Percentage lost') +
  xlab('N species protected') +
  scale_color_manual(values = c('blue', 'darkorange', 'darkgrey')) +
  scale_x_continuous(trans = 'log10') +
  theme(legend.title = element_blank())
g8
```

The same patterns hold under the pessimistic scenario.

### 10.3 Stochastic prioritisation efficiency

To formally test whether differences between prioritisation strategies are statistically significant, we re-run the simulations drawing the number of protected species at random from a uniform distribution (1,000 iterations). 

We also include a fourth treatment: protecting species in order of endangerment score only, with ties broken by random shuffling of the data frame before sorting.

```{r}
EDGE_species  <- df$Species[order(df$EDGE,  decreasing = T)]
FuDGE_species <- df$Species[order(df$FuDGE, decreasing = T)]

PDlost <- vector()
FDlost <- vector()
df$extinction_prob[which(df$assessment == 'Least Concern')]         <- 0.0017
df$extinction_prob[which(df$assessment == 'Near Threatened')]       <- 0.0141
df$extinction_prob[which(df$assessment == 'Vulnerable')]            <- 0.1
df$extinction_prob[which(df$assessment == 'Endangered')]            <- 0.6723
df$extinction_prob[which(df$assessment == 'Critically Endangered')] <- 0.999

arksize <- round(runif(1000, 0, length(EDGE_species)))

for(i in 1:1000) {
  df_clone <- df
  df_clone$extinction_prob[which(df_clone$Species %in% EDGE_species[1:arksize[i]])] <- 0
  keepers <- df_clone$Species[which(unlist((lapply(df_clone$extinction_prob, extinction_roll))) == 'survived')]
  phy_extinct <- keep.tip(phy, keepers)
  PDlost <- c(PDlost, 1 - (sum(phy_extinct$edge.length) / sum(phy$edge.length)))
  fun_extinct <- keep.tip(fun_tree, keepers)
  FDlost <- c(FDlost, 1 - (sum(fun_extinct$edge.length) / sum(fun_tree$edge.length)))
}

output <- tibble(Runs = 1:1000, PD_EDGE = PDlost, FD_EDGE = FDlost)
tail(output)

PDlost <- vector()
FDlost <- vector()

for(i in 1:1000) {
  df_clone <- df
  df_clone$extinction_prob[which(df_clone$Species %in% FuDGE_species[1:arksize[i]])] <- 0
  keepers <- df_clone$Species[which(unlist((lapply(df_clone$extinction_prob, extinction_roll))) == 'survived')]
  phy_extinct <- keep.tip(phy, keepers)
  PDlost <- c(PDlost, 1 - (sum(phy_extinct$edge.length) / sum(phy$edge.length)))
  fun_extinct <- keep.tip(fun_tree, keepers)
  FDlost <- c(FDlost, 1 - (sum(fun_extinct$edge.length) / sum(fun_tree$edge.length)))
}

output$PD_FuDGE <- PDlost
output$FD_FuDGE <- FDlost

random_species <- sample(df$Species, replace = F)
PDlost <- vector()
FDlost <- vector()

for(i in 1:1000) {
  df_clone <- df
  df_clone$extinction_prob[which(df_clone$Species %in% random_species[1:arksize[i]])] <- 0
  keepers <- df_clone$Species[which(unlist((lapply(df_clone$extinction_prob, extinction_roll))) == 'survived')]
  phy_extinct <- keep.tip(phy, keepers)
  PDlost <- c(PDlost, 1 - (sum(phy_extinct$edge.length) / sum(phy$edge.length)))
  fun_extinct <- keep.tip(fun_tree, keepers)
  FDlost <- c(FDlost, 1 - (sum(fun_extinct$edge.length) / sum(fun_tree$edge.length)))
}

output$PD_Rnd <- PDlost
output$FD_Rnd <- FDlost

PDlost <- vector()
FDlost <- vector()

for(i in 1:1000) {
  df_endangered <- df[sample(1:nrow(df), nrow(df), replace = F),]
  endangered_species <- df_endangered$Species[order(df_endangered$GE, decreasing = T)]
  df_clone <- df
  df_clone$extinction_prob[which(df_clone$Species %in% endangered_species[1:arksize[i]])] <- 0
  keepers <- df_clone$Species[which(unlist((lapply(df_clone$extinction_prob, extinction_roll))) == 'survived')]
  phy_extinct <- keep.tip(phy, keepers)
  PDlost <- c(PDlost, 1 - (sum(phy_extinct$edge.length) / sum(phy$edge.length)))
  fun_extinct <- keep.tip(fun_tree, keepers)
  FDlost <- c(FDlost, 1 - (sum(fun_extinct$edge.length) / sum(fun_tree$edge.length)))
}

output$PD_End <- PDlost
output$FD_End <- FDlost

output <- gather(output, index, loss, PD_EDGE:FD_End)

output$species_saved <- rep(arksize, 8)

output$treatment <- 'EDGE'
output$treatment[grep('_FuDGE', output$index)] <- 'ecoDGE'
output$index <- gsub('_EDGE',  '', output$index)
output$index <- gsub('_FuDGE', '', output$index)
output$treatment[grep('_Rnd', output$index)] <- 'Rnd'
output$index <- gsub('_Rnd', '', output$index)
output$treatment[grep('_End', output$index)] <- 'End'
output$index <- gsub('_End', '', output$index)
```

```{r}
head(output)
```

```{r}
gg_scatter <- ggplot(output, aes(x = species_saved, y = (loss*100) + 0.001, col = treatment)) +
  scale_y_continuous(trans = 'log10') +
  scale_color_manual(values = c('#3f6faf', '#f49843', '#b43b4b', 'darkgrey'), name = "") +
  ylab('Percentage lost') +
  xlab('N species protected') +
  ggtitle('Optimistic scenario') +
  geom_smooth() +
  theme_classic()

ggsave("~/Library/CloudStorage/Dropbox/FuDGE/Figures/recovery_PD_FD.tiff", gg_scatter + facet_wrap(~ index), width = 7.5, height = 5)
gg_scatter + facet_wrap(~ index, ncol = 1, nrow = 2)
```

ecoDGE consistently outperforms EDGE at protecting FD, while EDGE only outperforms ecoDGE for PD when protecting very large numbers of species (~4,000+). For realistic intervention scales, protecting the most endangered species appears similarly effective for PD.

```{r}
gg_scatter + facet_wrap(~ index) + xlim(0, 2000)
```

```{r}
gg_scatter + facet_wrap(~ index) + xlim(0, 500) + scale_y_continuous() +
  geom_vline(xintercept = 100, lty = 2)
```

At the scale relevant to existing EDGE lists (top 100 species), ecoDGE performs best for FD and EDGE performs best for PD, though the gains over a simple endangerment-based approach are modest.

### 10.4 Pessimistic scenario stochastic runs

```{r}
output_opt <- output

EDGE_species  <- df$Species[order(df$EDGE,  decreasing = T)]
FuDGE_species <- df$Species[order(df$FuDGE, decreasing = T)]

df$extinction_prob[which(df$assessment == 'Least Concern')]   <- 0.01269
df$extinction_prob[which(df$assessment == 'Near Threatened')] <- 0.09
df$extinction_prob[which(df$assessment == 'Vulnerable')]      <- 0.605
df$extinction_prob[which(df$assessment == 'Endangered')]      <- 0.8991

PDlost <- vector()
FDlost <- vector()
arksize <- round(runif(1000, 0, length(EDGE_species)))

for(i in 1:1000) {
  df_clone <- df
  df_clone$extinction_prob[which(df_clone$Species %in% EDGE_species[1:arksize[i]])] <- 0
  keepers <- df_clone$Species[which(unlist((lapply(df_clone$extinction_prob, extinction_roll))) == 'survived')]
  phy_extinct <- keep.tip(phy, keepers)
  PDlost <- c(PDlost, 1 - (sum(phy_extinct$edge.length) / sum(phy$edge.length)))
  fun_extinct <- keep.tip(fun_tree, keepers)
  FDlost <- c(FDlost, 1 - (sum(fun_extinct$edge.length) / sum(fun_tree$edge.length)))
}

output <- tibble(Runs = 1:1000, PD_EDGE = PDlost, FD_EDGE = FDlost)
tail(output)

PDlost <- vector()
FDlost <- vector()

for(i in 1:1000) {
  df_clone <- df
  df_clone$extinction_prob[which(df_clone$Species %in% FuDGE_species[1:arksize[i]])] <- 0
  keepers <- df_clone$Species[which(unlist((lapply(df_clone$extinction_prob, extinction_roll))) == 'survived')]
  phy_extinct <- keep.tip(phy, keepers)
  PDlost <- c(PDlost, 1 - (sum(phy_extinct$edge.length) / sum(phy$edge.length)))
  fun_extinct <- keep.tip(fun_tree, keepers)
  FDlost <- c(FDlost, 1 - (sum(fun_extinct$edge.length) / sum(fun_tree$edge.length)))
}

output$PD_FuDGE <- PDlost
output$FD_FuDGE <- FDlost

random_species <- sample(df$Species, replace = F)
PDlost <- vector()
FDlost <- vector()

for(i in 1:1000) {
  df_clone <- df
  df_clone$extinction_prob[which(df_clone$Species %in% random_species[1:arksize[i]])] <- 0
  keepers <- df_clone$Species[which(unlist((lapply(df_clone$extinction_prob, extinction_roll))) == 'survived')]
  phy_extinct <- keep.tip(phy, keepers)
  PDlost <- c(PDlost, 1 - (sum(phy_extinct$edge.length) / sum(phy$edge.length)))
  fun_extinct <- keep.tip(fun_tree, keepers)
  FDlost <- c(FDlost, 1 - (sum(fun_extinct$edge.length) / sum(fun_tree$edge.length)))
}

output$PD_Rnd <- PDlost
output$FD_Rnd <- FDlost

PDlost <- vector()
FDlost <- vector()

for(i in 1:1000) {
  df_endangered <- df[sample(1:nrow(df), nrow(df), replace = F),]
  endangered_species <- df_endangered$Species[order(df_endangered$GE, decreasing = T)]
  df_clone <- df
  df_clone$extinction_prob[which(df_clone$Species %in% endangered_species[1:arksize[i]])] <- 0
  keepers <- df_clone$Species[which(unlist((lapply(df_clone$extinction_prob, extinction_roll))) == 'survived')]
  phy_extinct <- keep.tip(phy, keepers)
  PDlost <- c(PDlost, 1 - (sum(phy_extinct$edge.length) / sum(phy$edge.length)))
  fun_extinct <- keep.tip(fun_tree, keepers)
  FDlost <- c(FDlost, 1 - (sum(fun_extinct$edge.length) / sum(fun_tree$edge.length)))
}

output$PD_End <- PDlost
output$FD_End <- FDlost

output <- gather(output, index, loss, PD_EDGE:FD_End)

output$species_saved <- rep(arksize, 8)

output$treatment <- 'EDGE'
output$treatment[grep('_FuDGE', output$index)] <- 'ecoDGE'
output$index <- gsub('_EDGE',  '', output$index)
output$index <- gsub('_FuDGE', '', output$index)
output$treatment[grep('_Rnd', output$index)] <- 'Rnd'
output$index <- gsub('_Rnd', '', output$index)
output$treatment[grep('_End', output$index)] <- 'End'
output$index <- gsub('_End', '', output$index)

output_pesi <- output
```

### 10.5 Figure 3: Combined diversity-loss plot

```{r}
output_full <- rbind(
  output_opt  %>% mutate(Scenario = 'Optimistic'),
  output_pesi %>% mutate(Scenario = 'Pessimistic')
)

gg_scatter <- ggplot(output_full, aes(x = species_saved, y = (loss*100) + 0.001, col = treatment)) +
  scale_y_continuous(trans = 'log10') +
  scale_color_manual(values = c('#3f6faf', '#f49843', '#b43b4b', 'darkgrey'), name = "") +
  ylab('Percentage diversity lost') +
  xlab('N species protected') +
  geom_smooth() +
  theme_bw() +
  facet_wrap(Scenario ~ index)
gg_scatter
```

```{r}
F3 <- ggarrange(g6, gg_scatter, widths = c(1,2), labels = c('A', 'B'))
F3
ggsave('~/Dropbox/FuDGE/Figures/Figure3.pdf', plot = F3, width = 160, height = 90, units = 'mm', scale = 1.7)
```

```{r}
FigS2 <- gg_scatter + xlim(c(0,500)) + scale_y_continuous()
FigS2
ggsave('~/Dropbox/FuDGE/Figures/FigureS2.pdf', plot = FigS2, width = 120, height = 90, units = 'mm', scale = 1.7)
```


## 11. Figure 2: Index Score Distributions

We produce level plots showing how EDGE and FuDGE scores are distributed across the ED/FuD–GE space, overlaid with the actual species. We read in a pre-processed data file that retains all necessary columns.

```{r}
require(ggpubr)
require(viridis)

EDGE <- function(ED, GE) log(1 + ED) + GE * log(2)

df <- read.csv('bird-stuff.csv')
df <- df[order(df$EDGE, decreasing = T),]
df$chosen <- 'Non-Priority Species'
df$chosen[1:100] <- 'Priority Species'

data <- data.frame(ED = seq(0.2, 80, length.out = 100), GE = seq(-0.5, 4.5, length.out = 100))
data <- expand.grid(ED = data$ED, GE = data$GE)
data$EDGE <- with(data, EDGE(ED, GE))

g1 <- ggplot(data, aes(x = ED, y = GE, fill = EDGE, z = EDGE)) +
  geom_raster() +
  scale_fill_viridis(option = 'C', direction = 1, name = "Score") +
  scale_y_continuous(breaks = c(0:4), labels = c('LC', 'NT', 'VU', 'EN', 'CR')) +
  geom_jitter(data = df, height = 0.2, width = 0, size = 0.5,
              aes(x = ED, y = GE, col = chosen), show.legend = F) +
  scale_color_manual(values = c('lightgrey', 'black'), name = '') +
  theme_classic() +
  labs(title = 'EDGE') +
  theme(legend.position = 'bottom')

df <- df[order(df$FuDGE, decreasing = T),]
df$chosen <- 'Non-Priority'
df$chosen[1:100] <- 'Priority'
df$ecoDGE <- df$FuDGE

data <- data.frame(FuD = seq(0.2, 80, length.out = 100), GE = seq(-0.5, 4.5, length.out = 100))
data <- expand.grid(FuD = data$FuD, GE = data$GE)
data$ecoDGE <- with(data, EDGE(FuD, GE))

g2 <- ggplot(data, aes(x = FuD, y = GE, fill = ecoDGE, z = ecoDGE)) +
  geom_raster() +
  scale_fill_viridis(option = 'C', direction = 1, name = "Score") +
  scale_y_continuous(breaks = c(0:4), labels = c('LC', 'NT', 'VU', 'EN', 'CR')) +
  geom_jitter(data = df, height = 0.2, width = 0, size = 0.5,
              aes(x = FuD, y = GE, col = chosen), show.legend = F) +
  scale_color_manual(values = c('lightgrey', 'black'), name = '') +
  theme_classic() +
  labs(title = 'ecoDGE') +
  theme(legend.position = 'bottom')

grid <- ggarrange(g1, g2, nrow = 2, ncol = 1, hjust = -1,
                  common.legend = T, legend = 'right')
grid
```

```{r}
heatmap_plot <- heatmap_plot +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = 'none') +
  coord_equal()
heatmap_plot
F2 <- ggarrange(heatmap_plot, grid, widths = c(1.3, 1), nrow = 1, ncol = 2, labels = c('A', 'B'))
ggsave('~/Dropbox/FuDGE/Figures/Figure2.pdf', plot = F2, width = 160, height = 90, units = 'mm', scale = 1.7)
```

